Supplemental Information from:\\ \textit{Reconstructing the last known movements of one of Nature's giants}

### 2018-10-15 manuscript/NatureComms/blue-whale-ms-oct18.tex
Nature Comms version with new pacific analyses

\newpage
\section{Main text}
Migratory species pose a particular challenge for conservation practitioners because their effective conservation relies on protection at many, often distant, sites\cite{runge2014conserving}.
Migratory species may also be particularly vulnerable to changes in climate or human use of the environment, as they are influenced by conditions in multiple locations across different parts of their life cycle\cite{robinson2009travelling}. 
Identifying threats to migratory species and developing effective conservation measures requires information on individual-level movements spanning multiple years, for both historical and present-day populations. 
Such data are scarce, especially for rare, wide-ranging and long-lived marine species\cite{ryan2013stable,hall2005stable,bailey2009behavioural}.

>>>>

\newpage
\section{Main text}
Migratory species pose a particular challenge for conservation practitioners because their effective conservation relies on protection at many, often distant, sites\cite{runge2014conserving}.
Migratory species may also be particularly vulnerable to changes in climate or human use of the environment, as they are influenced by conditions in multiple locations across different parts of their life cycle\cite{robinson2009travelling}. 
Identifying threats to migratory species, understanding species responses to change and developing effective conservation measures all require information on the movements of individual animals over multiple years, ideally for both historic and present-day populations. 
With the development of electronic tagging technology, studies of the distributions of animals have largely shifted from infrequent observations of many individuals with limited or no information about individual movement behavior, to frequent observations of a few individuals, with detailed information about individual movement (Holdo et al., 2013).  
Despite the tremendous advances made using direct telemetry devices, tags are expensive, with a relatively high failure rate. 
Data on individual level, multi-annual movements remain scarce, especially for rare, wide-ranging and long-lived marine species such as baleen whales (Mysticeti)\cite{ryan2013stable,hall2005stable,bailey2009behavioural}.

An alternative technique for investigating animal movements retrospectively is to use intrinsic biochemical information such as stable isotope compositions. 
Present-day populations can be studied with material collected in the field, while historic samples can be taken from museum collections; rich archives of behavioral information that are often under-utilised. 
The stable isotope composition of animal tissues reflects the isotopic composition of diet at the time and place of ingestion, integrated over the timescale of tissue growth. 
A large literature describes the use of stable isotope markers to link the composition of a tissue to a geographic source at a single point in time (Hobson and Wassenaar 2008). 
Relating the isotopic compositions of marine animal tissues to the likely location of tissue growth is complicated by a lack of knowledge of spatial variations in the isotopic composition of diet (the isotopic baseline). 
This uncertainty is compounded when multiple samples are taken through time in the same individual, as temporal variation in the isotopic baseline must also be considered explicitly. 
Consequently, relatively little attention has been given to the potential to infer individual movement histories by reconstructing time series of multiple discrete origins from sequentially sampled incrementally-grown tissues.

Here we introduce a new approach to infer location from the isotopic compositions of sequentially-sampled incrementally-grown tissues by drawing on newly developed models predicting spatio-temporal variation in phytoplankton stable carbon isotope composition at global and monthly resolution (Magozzi et al., 2017; Fig. S3), coupled to an agent-based model of whale movements (see Supplementary Methods; Table S1; Fig. S5).
We use our coupled models to simulate time series of carbon stable isotope compositions expected under different movement behaviours.
We then compare simulated and measured with the measured isotopic profiles to identify potential movement histories most consistent with the measured isotopic profile.

We apply our method to blue whales (\textit{Balaenoptera musculus}), the largest animal to have ever lived. 
Mysticete whales are characterized by the development of baleen, keratinous structures in the upper jaw used to filter food items from seawater. 
Baleen is ideal for stable isotope studies because keratin grows continuously through an individual’s life, and once laid down it is metabolically inert (Schell, Schell Hobson ref). 
Baleen plates therefore offer a continuous isotopic record of behavior typically reflecting multiple years of life of an individual whale. 
B, although baleen is worn away at the tips over time, so a baleen plate reflects the most recent years of life, and rarely records an individual’s entire lifespan. 
Among the mysticete whales, blue whales are a particularly attractive target for model-based isotope movement work as they have a consistent, low trophic level diet and feed continuously through the year, perhaps driven by high energetic costs of maintaining extreme muscle mass (Goldbogen et al 2015).
Similar to other large balenopterid whales, blue whales are generally assumed to conduct annual migrations between high and low latitudes
Mysticete whales are characterized by the development of baleen, keratinous structures in the upper jaw used to filter food items from seawater. Baleen is ideal for stable isotope studies because keratin grows continuously through an individual’s life, and once laid down it is metabolically inert (Schell, Schell Hobson ref). Baleen plates therefore offer a continuous isotopic record of behavior reflecting multiple years of life of an individual whale, although baleen is worn away at the tips over time, so rarely records an individual’s entire lifespan. Baleen stable isotope records have a long history of use as biochemical markers to infer movement patterns in whales (e.g. Schell, Schell and Hobson), but have yet to be applied to blue whales from the North Atlantic population. we are not aware of any examples where baleen stable isotope data have been used to directly inform spatial management.  
Relating marine animal tissue isotopic compositions to likely locations is complicated by a lack of knowledge of spatio-temporal variations in the isotopic composition of diet. Here we introduce a new approach to infer location from tissue isotope data drawing on newly developed models predicting spatio-temporal variation in phytoplankton stable carbon isotope composition at global and monthly resolution (Fig. S3), coupled to an agent-based model of whale movements (see Supplementary Methods; Table S1; Fig. S5).  We use our coupled model to simulate time series of carbon stable isotope compositions expected under different movement behaviours. We are then able to compare simulated and measured with the measured isotopic profiles to identify the most likely movement history.

We apply this method to blue whales (Balaenoptera musculus), the largest animal to have ever lived. Blue whales are an attractive target for isotope-based movement work as they have a consistent, low trophic level diet and feed continuously through the year. Isotope data can be further complicated by seasonal or individual variations in diet compositions.
Here we apply this method to baleen from the Although hunting for blue whale has been banned for over 50 years, blue whales are still listed as Endangered in the IUCN Red List, and expected recoveries following the cessation of active hunting may be impeded by other human activities such as modifications to the food web or more direct impacts of noise and shipping.   Developing an understanding of the nature and plasticity of individual level movements in populations of blue (and other baleen) whales would aid management decision making in the conservation of this endangered species (Irvine et al 2017).   
 (Hucke-Gaete et al 2018). As blue whales feed throughout the year to accommodate energetic costs, migration routes and feeding areas for blue whales are thought to be shaped by the year-round location of highly productive regions (Branch et al 2007). Recent studies tracking movements of individual blue whales primarily in the northeast Pacific have demonstrated high levels of among-individual variation in movement histories including suspending migration for opportunistic feeding and skipped migration (i.e. year round residency in either summer or wint
er areas). most satellite deployments on blue whales report movement data for periods of time under 6 months (Heide-Jørgensen et al., 2001; Silva et al 2013; Bailey et al., 2009; Lesage et al 2017; Irvine et al 2017), and are unable to show individual movements between summer and winter feeding or breeding areas. T.

The NorthEast Pacific population of blue whales is perhaps the best studied in terms of individual level movements with a recent history of deployment of over 177 satellite tags (Mate et al 2007; Irvine et al 2017). 
 Similar to other large balenopterid whales, blue whales are generally assumed to conduct annual migrations between high and low latitudes. Blue whales, however are thought to feed throughout the year, perhaps driven by high energetic costs of maintaining extreme muscle mass (Goldbogen et al 2015). Migration routes and feeding areas are therefore thought to be shaped by the year-round location of highly productive regions (Branch et al 2007). The northeast Pacific population of blue whales is perhaps the best studied in terms of individual level movements with a recent history of deployment of over 177 satellite tags (Mate et al 2007; Irvine et al 2017). In the nNortheast Pacific, satellite tags show individuals occupying most of the exclusive economic zone (EEZ) and into the Gulf of Alaska (Irvine et al 2017 +) with greatest overlapping areas in the southern half of the EEZ, along the continental slope and shelf region between the California Channel Islands and Cape Mendocino. Whales were present in the EEZ throughout the year, but were mainly present between July and November. Whales tended to occupy more northerly portions of the range later in the year before departing the EEZ presumably for winter feeding grounds in late autumn. Winter feeding is less well characterized, but winter feeding grounds are known in the Gulf of California and in the eastern tropical Pacific, specifically the Costa Rica Dome  where production is maintained by seasonal upwelling (REFS). However, most satellite deployments on blue whales report movement data for periods of time under 6 months (Heide-Jørgensen et al., 2001; Silva et al 2013; Bailey et al., 2009; Lesage et al 2017; Irvine et al 2017), and are unable to show individual movements between summer and winter feeding or breeding areas. The longest period of continuous monitoring via satellite tagging of an individual blue whale that we are aware of in the northeast Pacific is 504 days (Irvine et al 2017) and in the north Atlantic is 177 days (Lesage et al 2017).

The blue whale (Balaenoptera musculus), the largest animal to have ever lived, is one such migratory marine species. Although hunting for blue whale has been banned for over 50 years, blue whales are still listed as Endangered in the IUCN Red List, and expected recoveries following the cessation of active hunting may be impeded by other human activities such as modifications to the food web or more direct impacts of noise and shipping.   Developing an understanding of the nature and plasticity of individual level movements in populations of blue (and other baleen) whales would aid management decision making in the conservation of this endangered species (Irvine et al 2017).  The northeast Atlantic population of blue whales was the first large whale population to be systematically hunted with explosive harpoons, and while the population was probably relatively small before hunting, now around 1000 individuals are estimated to be in the northeast Atlantic in summer, mainly distributed around Iceland (Pike, Víkingsson, Gunnlaugsson, & Øien, 2009). The small population size and oceanic habit makes blue whale movements extremely hard to study.

The NorthEast Pacific population of blue whales is perhaps the best studied in terms of individual level movements with a recent history of deployment of over 177 satellite tags (Mate et al 2007; Irvine et al 2017).   Similar to other large balenopterid whales, blue whales are generally assumed to conduct annual migrations between high and low latitudes. Blue whales, however are thought to feed throughout the year, perhaps driven by high energetic costs of maintaining extreme muscle mass (Goldbogen et al 2015). Migration routes and feeding areas are therefore thought to be shaped by the year-round location of highly productive regions (Branch et al 2007). In the Northeast Pacific satellite tags show individuals occupying most of the exclusive economic zone (EEZ) and into the Gulf of Alaska (Irvine et al 2017 +) with greatest overlapping areas in the southern half of the EEZ, along the continental slope and shelf region between the California Channel Islands and Cape Mendocino. Whales were present in the EEZ throughout the year, but were mainly present between July and November. Whales tended to occupy more northerly portions of the range later in the year before departing the EEZ presumably for winter feeding grounds in late autumn. Winter feeding is less well characterized, but winter feeding grounds are known in the Gulf of California and in the eastern tropical Pacific, specifically the Costa Rica Dome  where production is maintained by seasonal upwelling (REFS).

A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggests that blue whales in the northeast Atlantic also track regions of high production throughout the year, at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde Islands (M.E Baines & Reichelt, 2014). Northward migration in the spring may occur along mid Atlantic corridors with peak sightings in the Azores around April-May (Silva et al 2014). Summer feeding appears to occur in northerly latitudes around Iceland, and historically in the Norwegian and Barents Seas (Pike, Víkingsson, Gunnlaugsson, & Øien, 2009). Blue whales are frequently detected in waters to the west of the UK, with peak acoustic detections occurring between November and December, in southerly migrating animals (M. E Baines, Reichelt, & Griffin, 2017; Charif & Clark, 2009; Reeves, Smith, Josephson, Clapham, & Woolmer, 2004; Visser, Hartman, Pierce, Valavanis, & Huisman, 2011). To our knowledge one photographic identification has matched summer and winter locations of a single north-east Atlantic blue whale with sightings in Iceland and Mauritania (REF).

Inferences concerning movements of blue whales in general and in the northeast Atlantic in particular are therefore largely drawn from disparate information sources and rarely detail movements of individual animals over timescales of a year or more. It is particularly difficult to establish the degree of connectivity between summer and winter feeding areas, and the level of temporal consistency in movement behaviour within individuals. Information on historical movement patterns of individual blue whales is important for understanding the drivers of whale declines, and potential for human-whale conflicts as populations recover.

Here we infer individual level high- resolution movement records for seven blue whales for whom baleen stable isotope data were available. Isotope data forro 6 whales stranding on the west coast of the USA or Mexico were reported in Busquetts- Vass et al 2016, and we report new isotopic data from a whale stranding off the coast of Wexford Ireland in March 1891, and currently on display in the Natural History Museum London’s (NHM) blue whale ‘Hope’.   Hope The NHM whale (Cat No XXXX, unofficially christened ‘Hope”) stranded in Wexford Ireland in March 1891 and is a female estimated to be a young adult at least 15 years at  death, suggesting that her. Bbaleen from the NHM whale will therefore yields information about migratory behaviors in north Atlantic blue whales around the peak of industrial whaling.

### 2018-10-15 manuscript/NatureComms/blue-whale-supplementary-oct18.tex
Initial commit of new Oct 2018 version
Nature Comms version with new pacific analyses

% Preamble
\documentclass[a4paper, 12pt]{article}

\usepackage[osf]{mathpazo} % palatino
%\usepackage{ms1}            % load the template
%\usepackage[round]{natbib} % author-year citations
\usepackage[superscript,biblabel]{cite} % for superscript citations
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{parskip} 
\usepackage{amsmath}
\usepackage{textcomp} % for parts per mille symbol  
\usepackage{pdflscape} % landscape   
\pagenumbering{arabic}

\linespread{1.66}

@@

\section*{Supplemental Results}

@@

% References
\newpage
\bibliographystyle{currentbiol}
\bibliography{blue-whale-abbreviations}

>>>>

% Preamble
\documentclass[a4paper,10pt]{article}

%\usepackage[osf]{mathpazo} % palatino
%\usepackage{ms1}            % load the template
%\usepackage[round]{natbib} % author-year citations
\usepackage[superscript,biblabel]{cite} % for superscript citations
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{parskip} 
\usepackage{amsmath}
\usepackage{textcomp} % for parts per mille symbol  
\usepackage{pdflscape} % landscape   
\pagenumbering{arabic}

% Weird Current Biology rules about times and single line spacing
\linespread{1}
\usepackage{times}

@@

\section*{Supplemental Results: NHM blue whale}

@@

\section*{Supplemental Results: Pacific blue whales from Busquets-Vass et al. 2017}

\begin{landscape}
    \begin{figure}[!htbp]
    \centering
      \includegraphics[width=\linewidth]{figures/top25.moveWhaleA_TL2.eps}
      \caption{Simulated locations by month taken from the top 25\% best fitting migratory movement model simulations for Pacific blue whale A.}
      \label{figwhaleA}
  \end{figure}
\end{landscape}

\begin{landscape}
    \begin{figure}[!htbp]
    \centering
      \includegraphics[width=\linewidth]{figures/top25.moveWhaleB_TL2.eps}
      \caption{Simulated locations by month taken from the top 25\% best fitting migratory movement model simulations for Pacific blue whale B.}
      \label{figwhaleB}
  \end{figure}
\end{landscape}

\begin{landscape}
    \begin{figure}[!htbp]
    \centering
      \includegraphics[width=\linewidth]{figures/top25.moveWhaleC_TL2.eps}
      \caption{Simulated locations by month taken from the top 25\% best fitting migratory movement model simulations for Pacific blue whale C.}
      \label{figwhaleC}
  \end{figure}
\end{landscape}

\begin{landscape}
    \begin{figure}[!htbp]
    \centering
      \includegraphics[width=\linewidth]{figures/top25.moveWhaleD_TL2.eps}
      \caption{Simulated locations by month taken from the top 25\% best fitting migratory movement model simulations for Pacific blue whale D.}
      \label{figwhaleD}
  \end{figure}
\end{landscape}

\begin{landscape}
    \begin{figure}[!htbp]
    \centering
      \includegraphics[width=\linewidth]{figures/top10.moveWhaleE_TL2.eps}
      \caption{Simulated locations by month taken from the top 10\% best fitting migratory movement model simulations for Pacific blue whale E.}
      \label{figwhaleE}
  \end{figure}
\end{landscape}

\begin{landscape}
    \begin{figure}[!htbp]
    \centering
      \includegraphics[width=\linewidth]{figures/top25.moveWhaleF_TL2.eps}
      \caption{Simulated locations by month taken from the top 25\% best fitting migratory movement model simulations for Pacific blue whale F.}
      \label{figwhaleF}
  \end{figure}
\end{landscape}

% References
\newpage
\bibliographystyle{nature}
\bibliography{blue-whale-abbreviations}
